---
title: "Assignment 2: Coding Basics"
author: Tasha Griffiths
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
hundredbyfour <- seq(1, 100, 4) 
#creating a value named 'hundredbyfour' that captures a sequence of numbers from 1 to 100 by 4

#2. 
mean(hundredbyfour) 
median(hundredbyfour)
#using a function to pull the mean and median of the pre-set named value 'hundredbyfour'

#3.
mean(hundredbyfour) > median(hundredbyfour)
#using a function to pull the mean of 'hundredbyfour' and then using basic math symbols to see if the mean value is greater than the median 'hundredbyfour'

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
Student_First_Name <- c('Tasha', 'Erick', 'Izzy', 'Meilin') 
#Vector-Class is Character

Test_Scores <- c(46, 71, 92, 100)
#Vector-Class is Numeric

Student_Passed <- c(FALSE, TRUE, TRUE, TRUE)
#Vector-CLass is logic

df_Student_Test1_Results <- as.data.frame(Student_First_Name)
df_Student_Test1_Results <- cbind(Student_First_Name, Test_Scores, Student_Passed)
df_Student_Test1_Results
#creating dataframe from main value of name, then combining 3 other values to make columns 

row.names(df_Student_Test1_Results)
is.data.frame(df_Student_Test1_Results)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Dataframes pull together different types or classes of vectors/values (characters, numbers, logic), however a matrix can only combine a single vecto-class.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
Is_Student_Passing <- function(x) {
ifelse(x>50, Pass, Fail)
}

Is_Student_Passing(df_Student_Test1_Results$Test_Scores )
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
