---
title: "Assignment 3: Data Exploration"
author: "Tasha Griffiths, Section #2"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}

#check working directory location
getwd()

#install package if not on system
#install.packages("tidyverse")

#load package for session
library(tidyverse)
library(formatR)

#load two datasets
Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: By looking at the ecotoxicology of neonicotinoids on insects we can better understand the toxic effects on humans. We can also understand if the insecticides are impacting more species than just insects, or if they are effective in killing insects. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: By understanding the litter in a forest, you can better understand impacts on soil health, tree growth and decline, and insect communities. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* Litter and woody debris are collected from both elevated and ground traps
* Mass measurements are made and debris sorted
* Locations of tower plots are selected randomly, in forest plots 40m X 40m
* 1-4 trap pairs per plot. Spacing requirements are set between plots, roads, streams, and buildings.

## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}

#basic summaries of dataset
class(Neonics)
summary(Neonics)

#to find dimensions of dataset
length(Neonics)
dim(Neonics)

#to view dataset
colnames(Neonics)
str(Neonics)

```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}

summary(Neonics$Effect)

```

> Answer: Most common effects for the insecticide data are population- which can be due to researchers wanted to see if numbers of a particular insect are lowered by the chemical. Next is mortality, which can be to see if the chemical succeeded in killing the subject. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}

summary(Neonics$Species.Common.Name)

```

> Answer: 5 of the 6 of the most common insects studied are in the bee family which is an important group of pollinators. This group may be of interest because farmers may not want their insecticides to harm bee species who are important in pollinating their crops. The other common insect is the parasitic wasp, which is another helpful insect for farmers since they kill many other kinds of insect pests.  

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}

class(Neonics$Conc.1..Author.)
summary(Neonics$conc.1..Author.)

```

> Answer: Its not numeric because the data imported had both numbers, characters, and other text that was then imported with stringsasfactors which then switched the class to 'factor'.  



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 38) 

#using 38 for bins since there has been 38 years of data collected in the study. 


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}

ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 38)

#color added as a separate aestheic in the chain argument. 

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Most common test location is the lab, which makes sense since it is a more controled location where you can handle a lot of samples at one time. Next frequent is the natural field. You can see that natural field was higher than lab from 1990 through 2000, and then lab had a major peak - I imagine this difference is due to advances in technology over that 10 year time period, and over the next 20 years technology increases would lead to more and more lab based samples. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Neonics$Endpoint)) +
  geom_bar() +
theme(axis.text.x = element_text(angle = 70, hjust = 1))

#largest is NOEC, second is LOEC

```

> Answer: Most common endpoint is NOEC, which means 'No observable effect concentrations' used in aquatic systems and the next most common is LOEC which means 'No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test' in terrestrial. Both of these endpoints refer to no-effects of the insecticide.
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate) #its a factor
tail(Litter$collectDate) #used to see what the current year,month,date format is

#old code
#as.character(Litter$collectDate)
#convert to date
#Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y/%m/%d")
#class(Litter$collectDate) #conversion worked class is now a date but all data is N/A

# Feb 1 class note: problem was the slashes in y / m/ d - should have been dashes so that it matches how the data is in the column already

#install lubridate
library(lubridate)

#convert to date with year in first
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)

#which dates were sampled
unique(Litter$collectDate)

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

unique(Litter$namedLocation)

summary(Litter$namedLocation)

```

> Answer: There are 12 unique plots on Niwot Ridge. Summary is different from unique because summary 'summarizes' the number of times a unique name was used (provides statistics on those values) within the column while unique just pulls how many unqiue (individual) names are within the column, not how often they are used.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x = Litter$functionalGroup)) +
  geom_bar()


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

ggplot(Litter) +
  geom_boxplot(aes(x = Litter$functionalGroup, y = Litter$dryMass), bins = 50) 

ggplot(Litter) +
  geom_violin(aes(x = Litter$functionalGroup, y = Litter$dryMass), bins = 50) 


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Violin polots are useful when seeing if there are a lot of repetitions around a certain value, however in this case there is essentailly no repitiion so the violin plots are too thin to be useful. This means that the range of distrubution among quartiles is easier to determine from the boxplot. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles becuase there is more range and a wider box for that functional group. 