---
title: "Assignment 4: Data Wrangling"
author: Tasha Griffiths
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Monday, Feb 7 @ 7:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1
getwd()
#setwd("C:/Users/Tasha Griffiths/Documents/Duke Year 1/Spring 22 Classes/Environmental Data Analytics/GithubRepos/Environmental_Data_Analytics_2022")

#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
#install.packages(dplyr)
library(dplyr)

#load all 4 raw data files
EPAair_03_NC2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", 
                             stringsAsFactors = TRUE)
EPAair_03_NC2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", 
                             stringsAsFactors = TRUE)
EPAair_PM25_NC2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", 
                               stringsAsFactors = TRUE)
EPAair_PM25_NC2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", 
                               stringsAsFactors = TRUE)

#2
#basic exploration, repeat for all datasets
colnames(EPAair_03_NC2018)
head(EPAair_03_NC2018)
summary(EPAair_03_NC2018)
str(EPAair_03_NC2018)
dim(EPAair_03_NC2018)

```

## Wrangle individual datasets to create processed files.
3. Change date to a date object
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 check data class 
class(EPAair_03_NC2018$Date)
class(EPAair_03_NC2019$Date)
class(EPAair_PM25_NC2018$Date)
class(EPAair_PM25_NC2019$Date)

#Format all data file date columns as date
EPAair_03_NC2018$Date <- as.Date(EPAair_03_NC2018$Date, format = "%m/%d/%Y")
EPAair_03_NC2019$Date <- as.Date(EPAair_03_NC2019$Date, format = "%m/%d/%Y")
EPAair_PM25_NC2018$Date <- as.Date(EPAair_PM25_NC2018$Date, format = "%m/%d/%Y")
EPAair_PM25_NC2019$Date <- as.Date(EPAair_PM25_NC2019$Date, format = "%m/%d/%Y")


#4 select a subset of columns
EPAair_03_NC2018 <- select(EPAair_03_NC2018, Date, DAILY_AQI_VALUE, Site.Name, 
                           AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
                           SITE_LONGITUDE)

EPAair_03_NC2019 <- select(EPAair_03_NC2019, Date, DAILY_AQI_VALUE, Site.Name, 
                           AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
                           SITE_LONGITUDE)

EPAair_PM25_NC2018 <- select(EPAair_PM25_NC2018, Date, DAILY_AQI_VALUE, 
                             Site.Name, AQS_PARAMETER_DESC, COUNTY, 
                             SITE_LATITUDE, SITE_LONGITUDE)

EPAair_PM25_NC2019 <- select(EPAair_PM25_NC2019, Date, DAILY_AQI_VALUE, 
                             Site.Name, AQS_PARAMETER_DESC, COUNTY, 
                             SITE_LATITUDE, SITE_LONGITUDE)

#5 fill cells in MP25 data sets
EPAair_PM25_NC2018$AQS_PARAMETER_DESC <- 'PM2.5'

EPAair_PM25_NC2019$AQS_PARAMETER_DESC <- 'PM2.5'


#6 saved as data files as processed
write.csv(EPAair_03_NC2018, row.names = FALSE, 
          file = "./Data/Processed/EPAair_03_NC2018_processed.csv")

write.csv(EPAair_03_NC2019, row.names = FALSE, 
          file = "./Data/Processed/EPAair_03_NC2019_processed.csv")

write.csv(EPAair_PM25_NC2018, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")

write.csv(EPAair_PM25_NC2019, row.names = FALSE, 
          file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Filter records to include just the sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School". (The `intersect` function can figure out common factor levels if we didn't give you this list...)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC2122_Processed.csv"

```{r}
#7 combine all dataframes
EPAair_O3_2018and2019 <- full_join(EPAair_03_NC2018, EPAair_03_NC2019)

EPAair_PM25_2018and2019 <- full_join(EPAair_PM25_NC2018, EPAair_PM25_NC2019)

EPAair_O3_PM25_NC2122 <- full_join(EPAair_O3_2018and2019, EPAair_PM25_2018and2019)

#filter with intersect gives you all the things that are common intbetween the two data sets. So you can use 2018 data and 2019 data and intsect to see all the sites that are common between the two. 

#8 Filter dataset with pipe
EPAair_O3_PM25_NC2122_filtered <- EPAair_O3_PM25_NC2122 %>%
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
        "Hattie Avenue", "Clemmons Middle", "Mendenhall School", 
        "Frying Pan Mountain", "West Johnston Co.", "Garinger High School",
        "Castle Hayne", "Pitt Agri. Center", "Bryson City", 
        "Millbrook School")) %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarize(AQI_mean = mean(EPAair_O3_PM25_NC2122$DAILY_AQI_VALUE),
            latitude_mean = mean(EPAair_O3_PM25_NC2122$SITE_LATITUDE),
            longitude_mean = mean(EPAair_O3_PM25_NC2122$SITE_LONGITUDE)) %>%
  mutate(Month = month(Date)) %>%
  mutate(Year = year(Date))

dim(EPAair_O3_PM25_NC2122_filtered)

#9 spread ozone and PM2.5
EPAair_O3_PM25_NC2122_spread <- EPAair_O3_PM25_NC2122_filtered %>%
  pivot_wider(names_from = AQS_PARAMETER_DESC, values_from = AQI_mean)

#10
dim(EPAair_O3_PM25_NC2122_spread)

#11
write.csv(EPAair_O3_PM25_NC2122_spread, row.names = FALSE, 
          file ="./Data/Processed/EPAair_O3_PM25_NC2122_Processed.csv")

```


## Generate summary tables
12a. Use the split-apply-combine strategy to generate a summary data frame from your results from Step 9 above. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. 

12b. BONUS: Add a piped statement to 12a that removes rows where both mean ozone and mean PM2.5 have missing values. 

13. Call up the dimensions of the summary dataset.


```{r}
#12(a,b)
EPAair_O3_PM25_NC2122_summary <- EPAair_O3_PM25_NC2122_spread %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(AQI_mean_ozone = mean(Ozone),
            AQI_mean_PM2.5 = mean(PM2.5)) %>%
  filter(!is.na(AQI_mean_ozone & AQI_mean_PM2.5))

#filter(drop_na(AQI_mean_ozone & AQI_mean_PM2.5)) drop_na only works for one 
#column at a time not able to use two columns at once. 

#13
dim(EPAair_O3_PM25_NC2122_summary)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: drop_na will remove NA's within the column that you specify, but does not use a logic to check between two columns. To remove rows with NA's as true in both columns, we need to use a filter at the is.na function. The na.omit works by removing rows that have any NA's within them, so if an NA exists in ozone or PM2.5 it will be dropped. However, it doesn't check for NA's within both columns.  
