---
title: "Assignment 5: Data Visualization"
author: "Tasha Griffiths"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
           stringsAsFactors = TRUE)
Niwot.Ridge.litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
                               stringsAsFactors = TRUE)

class(Niwot.Ridge.litter$collectDate)
class(PeterPaul.chem.nutrients$sampledate)

Niwot.Ridge.litter$collectDate <- as.Date(Niwot.Ridge.litter$collectDate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")



#2 
class(Niwot.Ridge.litter$collectDate)
class(PeterPaul.chem.nutrients$sampledate)

Niwot.Ridge.litter$collectDate <- as.Date(Niwot.Ridge.litter$collectDate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")



```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3

mytheme <- theme_classic(base_size = 12) + 
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")

#scale_color_manual(values=c("olivedrab4", "steelblue3"))

theme_set(mytheme) #set theme for all subsequent plots

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4

library(ggplot2)

NTL.LTER.plot <- ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = po4, 
                                                      color = lakename)) + 
  geom_point() + ylim(0, 50) + #add points and limit outliers
  scale_color_manual(values=c("olivedrab4", "steelblue3")) + #separate colors for each lake
  facet_wrap('lakename') +  xlab("Total Phosphorus") + ylab("Phosphate") + #change axis labels to be more clear 
  geom_smooth(colour="black", method = lm) #adding linear best fit line
print(NTL.LTER.plot)



```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5

class(PeterPaul.chem.nutrients$month)
PeterPaul.chem.nutrients$month <- as.factor(PeterPaul.chem.nutrients$month)

#boxplot for a - temperature
boxplot.temp <- ggplot(PeterPaul.chem.nutrients, aes( x = month, 
                                                      y = temperature_C, )) + 
  geom_boxplot(aes(color = lakename))
print(boxplot.temp)
  
#boxplot for b - TP
boxplot.tp <- ggplot(PeterPaul.chem.nutrients, aes( x = month, 
                                                      y = tp_ug, )) + 
  geom_boxplot(aes(color = lakename))
print(boxplot.tp)

#boxplot for c - TN
boxplot.tn <- ggplot(PeterPaul.chem.nutrients, aes( x = month, 
                                                      y = tn_ug, )) + 
  geom_boxplot(aes(color = lakename))
print(boxplot.tn)

#cowplot that combines all 3 boxplots
cowplot.temp.tn.tp <- plot_grid(boxplot.temp, boxplot.tp, boxplot.tn, 
                                nrow = 1, align = 'h', align = rel_heights = c(1.25, 1))
print(cowplot.temp.tn.tp)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
