---
title: "Assignment 09: Data Scraping"
author: Tasha Griffiths
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1. check directory
getwd()


#2. load libraries
library(tidyverse)
library(lubridate)
library(viridis)
library(rvest)
library(dataRetrieval)
library(tidycensus)

# Set theme
mytheme <- theme_classic() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website}
#2
webpageNC_DEQ <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')
webpageNC_DEQ

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data}
#3
water.system.name <- webpageNC_DEQ %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water.system.name

pswid <- webpageNC_DEQ %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pswid
  
ownership <- webpageNC_DEQ %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership
  
max.withdrawals.mgd <- webpageNC_DEQ %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max.withdrawals.mgd

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data}
#4 convert max withdrawls to dataframe

#need to pull month information from web table
max.withdrawls.month <- webpageNC_DEQ %>% 
  html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% 
  html_text()
max.withdrawals.month

#create a year and date variable to add to dataframe
max.withdrawls.year <- 2020
Date <- as.Date(my(paste(max.withdrawls.month, "-", max.withdrawls.year)))
class (Date)

#create the dataframe
the_df <- data.frame(
"Water System Name" = water.system.name,
"PSWID" = pswid,
"Ownership" = ownership,
"Max Withdrawls Total" = as.numeric(max.withdrawals.mgd),
"Max Withdrawls Month" = max.withdrawls.month,
"Max Withdrawls Year" = max.withdrawls.year,
"Date" = as.Date(my(paste(max.withdrawls.month, "-", max.withdrawls.year)))
)


#5 Plot couldn't get a line to work? maybe bar chart better?
ggplot(the_df,aes(x=max.withdrawls.month,y=max.withdrawals.mgd)) + 
  geom_point() +
  labs(title = paste("2020 Max Withdrawls for", water.system.name),
       subtitle = paste("PSWID", pswid),
       y="Max Withdrawls",
       x="Month")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function}
#6. create a function to scrape data for any PSWID and Year

Withdrawls.scrape.function <- function(any_year, pswid_number){
  #fetch website
  the_url <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid='
                    ,pswid_number, '&year=',any_year))
  print(the_url)
  
  #scrape data
  water.system.name.new <- "div+ table tr:nth-child(1) td:nth-child(2)"
  pswid.new <- "td tr:nth-child(1) td:nth-child(5)"
  ownership.new <- "div+ table tr:nth-child(2) td:nth-child(4)"
  max.withdrawals.mgd.new <- "th~ td+ td"
  max.withdrawls.month.new <- ".fancy-table:nth-child(31) tr+ tr th"
  max.withdrawls.year.new <- any_year
  
  
  web.system.name <- the_url %>% html_nodes(water.system.name.new) %>% html_text()
  web.pswid <- the_url %>% html_nodes(pswid.new) %>% html_text()
  web.ownership <- the_url %>% html_nodes(ownership.new) %>% html_text()
  web.max.withdrawls <- the_url %>% html_nodes(max.withdrawals.mgd.new) %>% html_text()
  web.withdrawls.month <- the_url %>% html_nodes(max.withdrawls.month.new) %>% html_text()
 
  #convert to dataframe
  new.dataframe <- data.frame("Water System Name" = web.system.name,
                              "PSWID" = web.pswid,
                              "Ownership" = web.ownership,
                              "Max_Withdrawls_Total" = as.numeric(web.max.withdrawls),
                              "Max_Withdrawls_Month" = web.withdrawls.month,
                              "Max_Withdrawls_Year" = max.withdrawls.year.new #,
                              #"Date" = as.Date(my(paste(web.max.withdrawls, "-",
                                                        #max.withdrawls.year.new)))
)
  
  #wrangle new dataframe
  new.dataframe <- new.dataframe %>% mutate("Date" = my(paste(Max_Withdrawls_Month, "-",
                                                        Max_Withdrawls_Year)))
  
  #show the dataframe
  return(new.dataframe)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 scape data for new year
durham.2015.withdrawls <- Withdrawls.scrape.function('2015','03-32-010')
view(durham.2015.withdrawls)

#plot 1
ggplot(durham.2015.withdrawls) +
  geom_point(aes(x=Date,y=Max_Withdrawls_Total)) +
  labs(title = paste("2020 Max Withdrawls for", water.system.name),
       subtitle = paste("PSWID", durham.2015.withdrawls$PSWID),
       y="Max Withdrawls",
       x="Month")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.Durham.2015.data}
#8 scape data for new location
asheville.2015.withdrawls <- Withdrawls.scrape.function('2015','01-11-010')
view(asheville.2015.withdrawls)

#plot


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r}
#9 scrape 2010 to 2019
any_year = seq(2010,2020)
pswid = '01-11-010'


ggplot(durham.2015.withdrawls) +
  geom_point(aes(x=max.withdrawls.month.new,y=max.withdrawals.mgd.new)) +
  labs(title = paste("2020 Max Withdrawls for", water.system.name.new),
       subtitle = paste("PSWID", pswid.new),
       y="Max Withdrawls",
       x="Month")

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?